---
title: "Omega-3 and Inflammatory Arthritis Analysis"
author: "Ryan_Gan"
date: "January 23, 2017"
output: html_document
---

## Purpose of Markdown Document

This document contains the code used to produce the results for the paper, *The association between omega-3 fatty acid biomarkers and inflammatory arthritis in an at-risk population positive for anti-citrullinated protein antibodies.*

Calling libraries used for data manipulation and analysis.
```{r libraries, message=F, warning=F}

library(tidyverse) # data wrangling 
library(broom) # analysis object to tidy dataframe
library(survival) # time-varying model

```

## Baseline Analysis

Code and results for the cross-sectional analysis of baseline inflammatory arthritis (IA) prevalence and the association with omega-3 fatty acid (n-3 FA) biomarker, n-3 FA percentage of red blood cells (n-3 FA % in RBCs).

Reading baseline dataframe and selecting only 47 of the 657 variables that are relevant for the baseline analyses. 

### Note: I have some missing data for packyears and supplement use that I believe Elizabeth imputed based on subjects' subsequent records. Ask for an updated dataframe.

```{r read baseline dataframe, message=F, warning=F}

# reading baseline data
baseline_df <- read_csv("ia_baseline_analysis_df.csv") %>% 
  # limiting dataframe of 657 variables to relevant study variables
  select(ID, DOVisit, LABID, visit_num, Age, Gender, Sex, Race, Race_bi, Educ, 
         Inc, BMI, Packyears_new, SE_num, Omega3_bi, Multivits_bi, CCP3, CCP31, 
         CCP2, RFneph, CRPOutcome, total_n3, total_n6, ala, epa, dha, dpa, 
         epa_dha, la, gla, ara, f1perc, f2perc, stnd_ala, stnd_epa, stnd_dha, 
         stnd_dpa, stnd_epa_dha, stnd_total_n3, stnd_la, stnd_gla, stnd_ara, 
         stnd_total_n6, IA_base) %>% 
  # creating binary variables
  mutate(age50 = ifelse(Age >= 50, 1, 0), # age >= 50
         py10 = ifelse(Packyears_new > 10, 1, 0), # pack years more than 10 years
         crp = ifelse(CRPOutcome == "NN confirmed pos", 1, 0)) # crp positivity

```

Creating Table 1 of descriptive characteristics by baseline IA status. I applied Fisher's exact test to all categorical variables.


```{r table 1, message=F, warning=F}

# create empty dataframe of table 1
table1_matrix <- matrix(nrow = 14, ncol = 6, byrow = T)
# define column names for matrix
colnames(table1_matrix) <- c("variable", "ia_yes_n", "ia_yes_perc",
                             "ia_no_n", "ia_no_perc", "p_val")

# fill column 1 with variable names
table1_matrix[,1] <- c("n", "mean_age", "age50_perc", "female_perc", "nhw_perc",
  "ed_gt_hs", "income_gt_40k", "pack_years_gt_10", "shared_epitope_pos", 
  "n3_supp_use", "multi_vit_use", "rf_pos", "crp_pos", "ccp2_pos")

head(table1_matrix)

# calculate n
tab1_n_df <- baseline_df %>% 
  group_by(IA_base) %>% # group by IA status
  na.omit() %>% # removing missing
  summarise(n = n(), mean_age = n(), age50_n = sum(age50),
    female_n = sum(Sex), nhw_n = sum(Race_bi), ed_gt_hs_n = sum(Educ),
    income_gt_40k_n = sum(Inc), pack_years_gt_10_n = sum(py10),
    shared_epitope_pos_n = sum(SE_num), n3_supp_use_n = sum(Omega3_bi),
    multi_vit_use_n = sum(Multivits_bi), rf_pos_n = sum(RFneph),
    crp_pos_n = sum(crp), ccp2_pos_n = sum(CCP2)) 


# calculate percentage
tab1_perc_df <- baseline_df %>% 
  group_by(IA_base) %>% # group by IA status
  na.omit() %>% # removing missing
  summarise(n = n(), mean_age = mean(Age), age50_perc = sum(age50)/n,
    female_perc = sum(Sex)/n, nhw_perc = sum(Race_bi)/n, ed_gt_hs = sum(Educ)/n,
    income_gt_40k = sum(Inc)/n, pack_years_gt_10 = sum(py10)/n,
    shared_epitope_pos = sum(SE_num)/n, n3_supp_use = sum(Omega3_bi)/n,
    multi_vit_use = sum(Multivits_bi)/n, rf_pos = sum(RFneph)/n,
    crp_pos = sum(crp)/n, ccp2_pos = sum(CCP2)/n) %>% 
  round(., 2) # round to two digits

# fill columns ----
# fill IA yes n column
table1_matrix[ ,2] <- as.numeric(as.vector(tab1_n_df[2, 2:15]))
# fill IA no n column
table1_matrix[ ,4] <- as.numeric(as.vector(tab1_n_df[1, 2:15]))
# fill IA yes percent
table1_matrix[ ,3] <- as.numeric(as.vector(tab1_perc_df[2, 2:15]))
# fill IA no percent
table1_matrix[ ,5] <- as.numeric(as.vector(tab1_perc_df[1, 2:15]))

# perform statistics and fill colum n 6 ----

# calculate p value for test of mean differences
age_p <- as.vector(tidy(lm(Age ~ IA_base, baseline_df)) %>% 
  filter(term == "IA_base") %>% 
  select(p.value) %>% 
  round(., 3))

# fill test of mean difference in age p value
table1_matrix[2, 6] <- as.numeric(age_p)

# create loop to calculate p values for binary variables

# create vector of variables to loop
var_loop <- c("age50", "Sex", "Race_bi", "Educ", "Inc", "py10", "SE_num", 
              "Omega3_bi", "Multivits_bi", "RFneph", "crp", "CCP2")


for(i in 1:length(var_loop)){

  # create tab
  tab <- xtabs(as.formula(paste0("~IA_base +", var_loop[i])),
               data = baseline_df)
  
  # perform fisher's exact test; will be approximate to chisq with 
  # large enough sample size

  fisher_p <- tidy(fisher.test(tab)) %>% select(p.value)

  # fill in matrix
  table1_matrix[i + 2, 6] <- round(as.numeric(fisher_p), 3)
  
}





```