---
title: "Omega-3 and Inflammatory Arthritis Analysis"
author: "Ryan_Gan"
date: "January 23, 2017"
output: html_document
---

## Purpose of Markdown Document

This document contains the code used to produce the results for the paper, *The association between omega-3 fatty acid biomarkers and inflammatory arthritis in an at-risk population positive for anti-citrullinated protein antibodies.*

Calling libraries used for data manipulation and analysis.
```{r libraries, echo=F, message=F, warning=F}

library(tidyverse) # data wrangling 
library(broom) # analysis object to tidy dataframe
library(survival) # time-varying model
library(knitr) # for kable function

```

## Baseline Visit Cross-Sectional Analysis

Code and results for the cross-sectional analysis of baseline inflammatory arthritis (IA) prevalence and the association with omega-3 fatty acid (n-3 FA) biomarker, n-3 FA percentage of red blood cells (n-3 FA % in RBCs).

### Data Import

Reading baseline dataframe Elizabeth provided. There are 52 observations with 56 variables. Elizabeth limited dataframe to relevant variables and imputed some missing variables based on prior observations on the same subject.

```{r read baseline dataframe, echo=F, message=F, warning=F}

# reading baseline data
baseline_df <- read_csv("baseline.csv") %>% 
  rename(visit_num = visit_)

```

### Descriptive Table 1 by IA Status (Table 1)

Creating Table 1 of descriptive characteristics by baseline IA status. First step is creating an empty matrix and calculating cell values for each variable.

```{r table 1, echo=F, message=F, warning=F, results='asis'}

# create empty dataframe of table 1
table1_matrix <- matrix(nrow = 16, ncol = 8, byrow = T)
# define column names for matrix
colnames(table1_matrix) <- c("variable", "ia_yes_n", "ia_yes_val", "ia_yes_parenth",
                             "ia_no_n", "ia_no_val", "ia_no_parenth", "p_val")

# fill column 1 with variable names
table1_matrix[,1] <- c("n", "Age (Mean)", "Age > 50 %", "Female %", 
  "Non-Hispanic White %","Education > HS %", "Income > 40k %", "Ever Smoke",
  "Current Smoker", "Pack Years > 10 %", "SE+ %", "n-3 FA Supplement %", 
  "Multivitamin %", "RF+ %", "CRP+ %", "CCP2+ %")

#Now I want to calculate test statistics for each row variable. 
# As age is a continous variable, I test the differences between IA positive and 
# IA negative at baseline using a t-test (in a linear model). For the 2x2 
# contingency table tests, I use Fisher's exact test as it should be approximate 
# to Persons chi-squared for sufficiently large sample sizes.

# calculat IA n and fill first row n
tab <- xtabs(~ ia_base,  data = baseline_df)
table1_matrix[1:2,2] <- tab[2]
table1_matrix[1:2,5] <- tab[1]

# fill in means and SDs for IA cases
table1_matrix[2,3] <- round(mean(baseline_df$age[baseline_df$ia_base==1]),1)
table1_matrix[2,4] <- round(sd(baseline_df$age[baseline_df$ia_base==1]),1)
# fill in means and SDs for IA no 
table1_matrix[2,6] <- round(mean(baseline_df$age[baseline_df$ia_base==0]),1)
table1_matrix[2,7] <- round(sd(baseline_df$age[baseline_df$ia_base==0]),1)
# calculate p val
# calculate p value for test of mean differences
age_p <- as.vector(tidy(lm(age ~ ia_base, baseline_df)) %>% 
  filter(term == "ia_base") %>% 
  select(p.value) %>% 
  round(., 3))
# fill test of mean difference in age p value
table1_matrix[2, 8] <- as.numeric(age_p)

# create loop to calculate p values for binary variables
# create vector of variables to loop
var_loop <- c("age_bi", "sex", "race_bi", "educ", "inc", "eversmoke",
              "cursmoke", "py10", "se_bi", "omega3_bi", "multivits_bi", 
              "rfneph", "crp", "ccp2")

for(i in 1:length(var_loop)){

  # create tab
  tab <- xtabs(as.formula(paste0("~ia_base +", var_loop[i])),
               data = baseline_df)
  #calculate and round proportion
  tab_prop <- round(prop.table(tab, 1),3)*100 
  
  # fill columns ----
  # n sample size 
  table1_matrix[i+2,2] <- tab[2,2] + tab[2,1]
  # fill IA yes columns
  table1_matrix[i+2,3] <- tab[2,2]
  # fill with proportion/percentage
  table1_matrix[i+2,4] <- tab_prop[2,2]
  
  # fill IA no columns
  # n same size
  table1_matrix[i+2,5] <- tab[1,2] + tab[1,1]
  # fill IA no but yes to variable column
  table1_matrix[i+2,6] <- tab[1,2]
  # fill with proportion/percentage
  table1_matrix[i+2,7] <- tab_prop[1,2]
  # perform fisher's exact test; will be approximate to chisq with 
  # large enough sample size
  fisher_p <- tidy(fisher.test(tab)) %>% select(p.value)
  # fill in matrix
  table1_matrix[i + 2, 8] <- round(as.numeric(fisher_p), 3)
  
}

# convert table 1 matrix to dataframe
table1_df <- data.frame(table1_matrix)

knitr::kable(table1_df, caption = paste0("Table 1: Demographic and descriptive",
  " characteristics of ACPA positive subjects by Inflammatory Arthritis", 
  " (IA) status at baseline"))

```

Table 1 suggests a significant difference in proportions in the IA yes vs IA no group for n-3 FA supplement use, RF positivity, and CRP positivity.

### Association Between n-3 FA % in RBCs and Prevalence of IA at Baseline Visit (Table 2)

The following code contains logistic regression models for each n-3 FA % in RBCs biomarker. n-3 FA % in RBC biomarkers are: ALA, EPA, DPA, DHA, EPA+DHA, and ALA+EPA+DPA+DHA = Total n-3 FA. n-3 FA % in RBC variables were standardized, where the standard deviation of each n-3 FA distribution was divide from the distribution to allow for a "1 SD increase"" interpreation across all n-3 FA variables. Adjustment variables were based on confounding variables for variables associated with IA either in the baseline visit in the cross-sectional analysis or the baseline visit in the incidence study. 

```{r logistic regression loop, echo=F, message=F, warning=F, results='asis'}
# create empty matrix for table
tab2_matrix <- matrix(nrow = 6, ncol = 5 )
# column names of tables
colnames(tab2_matrix) <- c("n3_FA_RBC", "OR", "Lower_95_CI",
                           "Upper_95_CI", "p_value") 
# variables to loop through
n3_vars <- c("sd_ALA_update", "sd_EPA_update", "sd_dpa_update", 
             "sd_DHA_update", "sd_epa_dha_update", "sd_total_n3_update")

# fill first column of matrix with n3 vars
tab2_matrix[,1] <- c("ALA", "EPA", "DPA", "DHA", "EPA+DHA",
                     "Total n-3 FA")

# Iteration of logistic regression through n-3 FAs
for(i in 1:length(n3_vars)){
  # general logistc regression; using tidy from broom to output 
  # coeffecients
  log_reg_mod <- tidy(glm(as.formula(paste0("ia_base ~", 
                    paste(n3_vars[i], "omega3_bi", "py10", "rfneph", 
                          "crp", "se_bi", sep = "+"))),
                    data = baseline_df, family = "binomial"(link="logit"))) 
  
  # calculate odds ratio and 95% CI
  odds_ratio <- log_reg_mod %>% filter(term == n3_vars[i]) %>% 
    # estimate lower and upper 95% CI bounds
    mutate(lower_bound = estimate - (1.96*std.error),
           upper_bound = estimate + (1.96*std.error)) %>% 
    select(term, estimate, lower_bound, upper_bound, p.value)

  # fill ith row of table 2
  tab2_matrix[i,2:4] <- round(as.numeric(exp(odds_ratio[1, 2:4])),2)
  tab2_matrix[i,5] <- round(as.numeric(odds_ratio[1,5]),2)

}

# create table of logistic values
# convert table 1 matrix to dataframe
table2_df <- as.data.frame(tab2_matrix)

knitr::kable(table2_df, caption = paste0("Table 2: Association between n-3",
  " fatty % in red blood cells (RBC) and inflammatory arthritis (IA) in ACPA",
  " positive subjects at baseline visit."))
```

Logistic regression models were all adjusted for n-3 fatty acid supplement use, pack years > 10, RF+, CRP+, and SE. The odds ratios represent the difference in odds of IA for a one standard deviation (SD) difference in the n-3 FA% in RBC. The SD for these variables are listed below: ALA: 0.15, EPA: 0.54, DPA: 0.46, DHA: 1.19, EPA and DHA: 1.58, Total-n3: 1.88.

Increasing DPA, EPA+DHA, and Total n-3 FAs were significantly associated with a lower prevalence of IA.

### Plot of Table 2

Figure representation of table 2. Extreme confidence intervals make the scale of this plot a little crazy.

```{r alternate figure 2, echo=F, message=F, warning=F, results='asis'}
# convert to numeric
or_df <- tab2_matrix %>% as_data_frame() %>% 
  mutate(OR = as.numeric(OR), Lower_95_CI = as.numeric(Lower_95_CI),
         Upper_95_CI = as.numeric(Upper_95_CI), p_value = as.numeric(p_value))


or_fig <- ggplot(or_df, aes(x = n3_FA_RBC, y = OR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Lower_95_CI, ymax = Upper_95_CI),
                width = 0.15) +
  scale_y_log10(breaks = c(0.01, 0.1, 0.25, 0.5, 0.75, 1, 2.5, 5)) +
  geom_hline(yintercept = 1, linetype=2) +
  ylab('Odds Ratio for 1 Standard Deviation Increase') +
  xlab('Omega-3 Fatty Acid') +
  # plot theme
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

print(or_fig)
```

## Incident IA Time-Varying Analysis

Fill in descriptions of time-varying approach to analyze incident IA.

### Import Time-Varying Dataframe

Read in time-varying incident IA dataframe created by Elizabeth..

```{r read incident dataframe, echo=F, message=F, warning=F}

# reading baseline data
timevarying_df <- read_csv("incident_timevary.csv") %>% 
  rename(visit_num = visit_) %>% 
  # code indicator of IA incident group at baseline
  mutate(incident_ia_base = ifelse(ia_cat == 2, 1, 0))

# make a dataframe of baseline characteristics for IA group
tv_base_df <- timevarying_df %>% filter(visit_num == 1)

```

### Descriptive Characteristics at Baseline Visit of IA Status (Table 3)

Calculate descriptive characteristics of incident IA dataframe by IA status at their baseline visit. I need to figure out how to handle correct income calculation in dataframe.

```{r table 3, echo=F, message=F, warning=F}

# create empty dataframe of table 1
table3_matrix <- matrix(nrow = 16, ncol = 8, byrow = T)
# define column names for matrix
colnames(table3_matrix) <- c("variable", "ia_yes_n", "ia_yes_val", "ia_yes_parenth",
                             "ia_no_n", "ia_no_val", "ia_no_parenth", "p_val")

# fill column 1 with variable names
table3_matrix[,1] <- c("n", "Age (Mean)", "Age > 50 %", "Female %", 
  "Non-Hispanic White %","Education > HS %", "Income > 40k %", "Ever Smoke",
  "Current Smoker", "Pack Years > 10 %", "SE+ %", "n-3 FA Supplement %", 
  "Multivitamin %", "RF+ %", "CRP+ %", "CCP2+ %")

#Now I want to calculate test statistics for each row variable. 
# As age is a continous variable, I test the differences between IA positive and 
# IA negative at baseline using a t-test (in a linear model). For the 2x2 
# contingency table tests, I use Fisher's exact test as it should be approximate 
# to Persons chi-squared for sufficiently large sample sizes.

# calculat IA n and fill first row n
tab <- xtabs(~ incident_ia_base,  data = tv_base_df)
table3_matrix[1:2,2] <- tab[2]
table3_matrix[1:2,5] <- tab[1]

# fill in means and SDs for IA cases
table3_matrix[2,3] <- round(mean(tv_base_df$age[tv_base_df$incident_ia_base==1]),1)
table3_matrix[2,4] <- round(sd(tv_base_df$age[tv_base_df$incident_ia_base==1]),1)
# fill in means and SDs for IA no 
table3_matrix[2,6] <- round(mean(tv_base_df$age[tv_base_df$incident_ia_base==0]),1)
table3_matrix[2,7] <- round(sd(tv_base_df$age[tv_base_df$incident_ia_base==0]),1)
# calculate p val
# calculate p value for test of mean differences
age_p <- as.vector(tidy(lm(age ~ incident_ia_base, tv_base_df)) %>% 
  filter(term == "incident_ia_base") %>% 
  select(p.value) %>% 
  round(., 3))
# fill test of mean difference in age p value
table3_matrix[2, 8] <- as.numeric(age_p)

# create loop to calculate p values for binary variables
# create vector of variables to loop
var_loop <- c("age_bi", "sex", "race_bi", "educ", "inc", "eversmoke", "cursmoke",
              "py10", "se_bi", "omega3_bi", "multivits_bi", "rfneph", "crp", "ccp2")

for(i in 1:length(var_loop)){

  # create tab
  tab <- xtabs(as.formula(paste0("~incident_ia_base +", var_loop[i])),
               data = tv_base_df)
  #calculate and round proportion
  tab_prop <- round(prop.table(tab, 1),3)*100 
  
  # fill columns ----
  # n sample size 
  table3_matrix[i+2,2] <- tab[2,2] + tab[2,1]
  # fill IA yes columns
  table3_matrix[i+2,3] <- tab[2,2]
  # fill with proportion/percentage
  table3_matrix[i+2,4] <- tab_prop[2,2]
  
  # fill IA no columns
  # n same size
  table3_matrix[i+2,5] <- tab[1,2] + tab[1,1]
  # fill IA no but yes to variable column
  table3_matrix[i+2,6] <- tab[1,2]
  # fill with proportion/percentage
  table3_matrix[i+2,7] <- tab_prop[1,2]
  # perform fisher's exact test; will be approximate to chisq with 
  # large enough sample size
  fisher_p <- tidy(fisher.test(tab)) %>% select(p.value)
  # fill in matrix
  table3_matrix[i + 2, 8] <- round(as.numeric(fisher_p), 3)
  
}

# convert table 1 matrix to dataframe
table3_df <- data.frame(table3_matrix)

knitr::kable(table3_df, caption = paste0("Table 3: Demographic and descriptive",
  " characteristics of ACPA positive subjects by incident Inflammatory Arthritis", 
  " (IA) status at baseline"))

```

### Association Between n-3 FA % in RBCs and Incident IA (Table 4)

Time-varying survival model that allows n-3 FA measure at each visit to contribute in the calculation of relative risk (hazard ratio; instantaenous risk of event). 

Have Elizabeth check to make sure I used the right n-3 FA standardized predictor. I have a feeling this might have changed when she updated the dataframe.

```{r timevary regression loop, echo=F, message=F, warning=F, results='asis'}
# create empty matrix for table
tab4_matrix <- matrix(nrow = 6, ncol = 5 )
# column names of tables
colnames(tab4_matrix) <- c("n3_FA_RBC", "HR", "Lower_95_CI",
                           "Upper_95_CI", "p_value") 
# variables to loop through
n3_vars <- c("sd_ala_update", "sd_epa_update", "sd_dpa_update", "sd_dha_update",
             "sd_epa_dha_update", "sd_total_n3_update")

# fill first column of matrix with n3 vars
tab4_matrix[,1] <- c("ALA", "EPA", "DPA", "DHA", "EPA+DHA",
                     "Total n-3 FA")

# Iteration of time varying model through n-3 FAs
for(i in 1:length(n3_vars)){

  # general logistc regression; using tidy from broom to output 
  # coeffecients
  tv_surv_mod <- tidy(coxph(as.formula(
    # formula for loop
    paste0("Surv(t1, t2, ia ==1) ~", 
    paste(n3_vars[i], "omega3_bi", "age_base_bi", "rfneph", 
    "crp", "se_bi", sep = "+"))),
    data = timevarying_df)) 
  

  # calculate hazard ratio and 95% CI
  hazard_ratio <- tv_surv_mod %>% filter(term == n3_vars[i]) %>% 
    # select terms to keep
    select(term, estimate, conf.low, conf.high, p.value)

  # fill ith row of table 2
  tab4_matrix[i,2:4] <- round(as.numeric(exp(hazard_ratio[1, 2:4])),2)
  tab4_matrix[i,5] <- round(as.numeric(hazard_ratio[1,5]),2)

}

# create table of logistic values
# convert table 1 matrix to dataframe
table4_df <- data.frame(tab4_matrix)


knitr::kable(table4_df, caption = paste0("Table 4: Association between increase in n-3",
  " fatty acid % in red blood cells (RBC) and incident inflammatory arthritis (IA) in ACPA",
  " positive subjects"))
```

### Plot of Table 4

Figure representation of table 4.

```{r hazard ratio plot, echo=F, message=F, warning=F, results='asis'}
# convert to numeric
hr_df <- tab4_matrix %>% as_data_frame() %>% 
  mutate(HR = as.numeric(HR), Lower_95_CI = as.numeric(Lower_95_CI),
         Upper_95_CI = as.numeric(Upper_95_CI), p_value = as.numeric(p_value))


hr_fig <- ggplot(hr_df, aes(x = n3_FA_RBC, y = HR)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = Lower_95_CI, ymax = Upper_95_CI),
                width = 0.15) +
  scale_y_log10(limits = c(0.1, 10), breaks = c(0.1, 0.25, 0.5, 0.75, 1, 1.75,
                                                2.5, 5.0, 10)) +
  geom_hline(yintercept = 1, linetype=2) +
  ylab('Hazard Ratio for 1 Standard Deviation Increase') +
  xlab('Omega-3 Fatty Acid') +
  # plot theme
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

print(hr_fig)
```


### Figure of n-3 FA Over Time

Kevin asked for a figure of n-3 FA levels over time. I'll see if this works. I think I'll go with a small-multiples figure by omega-3 FA.

First step is to get data in order for plotting over time. I need it in long form where there is a patient identifier, time in study, indicator if developed IA or no, omega of interest, and value.

The following figures show each subject's individual levels of n-3 FA over the study period, and is color-coded by whether they developed incident IA or not. The thicker lines are the fitted smooth splines for each group that sort of characterize the 'average' group effect over time. I do not show the error bars around the splines because there is way too much error. You would not be able to statistically tell if they are different from each other at any point. But maybe this helps to visualize trends over time.

#### Absolute n-3 Levels 

```{r n3 over time wrangle, echo=F, message=F, warning=F, results='asis'}

n3_time_plot <- timevarying_df %>%
  # limit to variables I need
  select(id, years_contributed, incident_ia_base, total_n3,
         ala, epa, dpa, dha, epa_dha) %>% 
  # use gather function to make a new var called omega
  gather(omega, perc_in_rbc, -years_contributed, -id, -incident_ia_base) %>% 
  mutate(id = as.factor(id), incident_ia_base = as.factor(incident_ia_base))  


# plot
n3_plot <-ggplot(n3_time_plot, aes(x=years_contributed, y = perc_in_rbc, 
  colour = incident_ia_base, group = interaction(id, incident_ia_base))) +
  geom_line(linetype = 2) +
  geom_smooth(aes(group = incident_ia_base), se = F) +
  facet_wrap(~omega,scales = "free") +
  theme_bw()

print(n3_plot)

```

These figures of the absolute levels don't really help. I think the individual levels jump around too much.

#### n-3 FA Change Visit to Visit

We could also look at change in n-3 FA levels from one visit to the next in a similar fashion.

```{r n3 chg plot, echo=F, message=F, warning=F, results='asis'}  
# change in n3 over time
n3_chg_plot <- timevarying_df %>%
  # limit to variables I need
  select(id, years_contributed, incident_ia_base, total_n3,
         ala, epa, dpa, dha, epa_dha) %>% 
  # calculate chang in n3
  group_by(id) %>% 
  mutate(total_n3_chg = total_n3 - lag(total_n3, order_by = id),
         ala_chg = ala - lag(ala, order_by = id),
         epa_chg = epa - lag(epa, order_by = id),
         dpa_chg = dpa - lag(dpa, order_by = id),
         dha_chg = dha - lag(dha, order_by = id),
         epa_dha_chg = epa_dha - lag(epa_dha, order_by = id)) %>% 
  ungroup() %>% 
  # limit to change variables
  select(-total_n3, -ala, -epa, -dpa, -dha, -epa_dha) %>% 
  # use gather function to make a new var called omega
  gather(omega, chg_n3, -years_contributed, -id, -incident_ia_base) %>% 
  mutate(id = as.factor(id), incident_ia_base = as.factor(incident_ia_base),
         chg_n3 = ifelse(is.na(chg_n3), 0, chg_n3)) 

# plot
n3_plot <-ggplot(n3_chg_plot, aes(x=years_contributed, y = chg_n3, 
  colour = incident_ia_base, group = interaction(id, incident_ia_base))) +
  geom_line(linetype = 2) +
  geom_smooth(aes(group = incident_ia_base), se = F) +
  facet_wrap(~omega,scales = "free") +
  theme_bw()

print(n3_plot)
```

These plots don't seem that helpful either, but it may suggest the people that didn't develop IA had relatively more stable n-3 FA levels over the time period. Spline for IA positive group may be a little questionable past 3 years since it's one subject driving the spline fit.

#### n-3 FA Change Since Baseline Visit

Last small multiples plot of n-3 FA change over time subtracts the baseline value of n-3 FA for each subject.

```{r n3 change since baseline vis, echo=F, message=F, warning=F, results='asis'}
# find baseline n3 vals
n3_base <- timevarying_df %>%
  # limit to variables I need
  select(id, total_n3, ala, epa, dpa, dha, epa_dha) %>% 
  # calculate chang in n3
  group_by(id) %>% 
  filter(row_number()==1) %>% 
  rename(base_tot_n3 = total_n3, base_ala = ala, base_epa = epa,
         base_dpa = dpa, base_dha = dha, base_epa_dha = epa_dha)

# dataset of differences since baseline visits
n3_chg_plot <- timevarying_df %>%
  # limit to variables I need
  select(id, years_contributed, incident_ia_base,
         total_n3, ala, epa, dpa, dha, epa_dha) %>% 
  # merge base values in
  right_join(n3_base, by = "id") %>% 
  # calculate chang in n3
  group_by(id) %>% 
  mutate(total_n3_chg = total_n3 - base_tot_n3,
         ala_chg = ala - base_ala,
         epa_chg = epa - base_epa,
         dpa_chg = dpa - base_dpa,
         dha_chg = dha - base_dha,
         epa_dha_chg = epa_dha - base_epa_dha) %>% 
  ungroup() %>% 
  # limit to change variables
  select(id, years_contributed, incident_ia_base, total_n3_chg,
         ala_chg, epa_chg, dpa_chg, dha_chg, epa_dha_chg) %>% 
  # use gather function to make a new var called omega
  gather(omega, chg_n3, -years_contributed, -id, -incident_ia_base) %>% 
  mutate(id = as.factor(id), incident_ia_base = as.factor(incident_ia_base),
         chg_n3 = ifelse(is.na(chg_n3), 0, chg_n3)) 

# plot
n3_plot <-ggplot(n3_chg_plot, aes(x=years_contributed, y = chg_n3, 
  colour = incident_ia_base, group = interaction(id, incident_ia_base))) +
  geom_line(linetype = 2) +
  geom_smooth(aes(group = incident_ia_base), se = F) +
  facet_wrap(~omega,scales = "free") +
  theme_bw()

print(n3_plot)

```

Again, I don't think there is a clear pattern here. You may conclude that those without incident IA may have relatively more stable n-3 FA levels over the study period. 


## Alternate Results Adjusting for Current Smoking Instead of PY10

Kevin is interested in controlling for the biologically/clinically more relevant smoking variable, "current smoker". 

### Baseline Results Adjusting for Current Smoker (Alt Table 2)

```{r alt logistic regression loop, echo=F, message=F, warning=F, results='asis'}
# create empty matrix for table
tab2_matrix <- matrix(nrow = 6, ncol = 5 )
# column names of tables
colnames(tab2_matrix) <- c("n3_FA_RBC", "OR", "Lower_95_CI",
                           "Upper_95_CI", "p_value") 
# variables to loop through
n3_vars <- c("sd_ALA_update", "sd_EPA_update", "sd_dpa_update", 
             "sd_DHA_update", "sd_epa_dha_update", "sd_total_n3_update")

# fill first column of matrix with n3 vars
tab2_matrix[,1] <- c("ALA", "EPA", "DPA", "DHA", "EPA+DHA",
                     "Total n-3 FA")

# Iteration of logistic regression through n-3 FAs
for(i in 1:length(n3_vars)){
  # general logistc regression; using tidy from broom to output 
  # coeffecients
  log_reg_mod <- tidy(glm(as.formula(paste0("ia_base ~", 
                    paste(n3_vars[i], "omega3_bi", "cursmoke", "rfneph", 
                          "crp", "se_bi", sep = "+"))),
                    data = baseline_df, family = "binomial"(link="logit"))) 
  
  # calculate odds ratio and 95% CI
  odds_ratio <- log_reg_mod %>% filter(term == n3_vars[i]) %>% 
    # estimate lower and upper 95% CI bounds
    mutate(lower_bound = estimate - (1.96*std.error),
           upper_bound = estimate + (1.96*std.error)) %>% 
    select(term, estimate, lower_bound, upper_bound, p.value)

  # fill ith row of table 2
  tab2_matrix[i,2:4] <- round(as.numeric(exp(odds_ratio[1, 2:4])),2)
  tab2_matrix[i,5] <- round(as.numeric(odds_ratio[1,5]),2)

}

# create table of logistic values
# convert table 1 matrix to dataframe
table2_df <- as.data.frame(tab2_matrix)

knitr::kable(table2_df, caption = paste0("Alt Table 2: Association between n-3",
  " fatty % in red blood cells (RBC) and inflammatory arthritis (IA) in ACPA",
  " positive subjects at baseline visit. Adjusting for current smoking rather than py10."))
```

Kevin, I thought a smoking variable was in our current incident model. I misunderstood the question you were answering. I'll adjust for current smoker in the baseline tables instead.